\Large Last Writer Slicing: Efficient Data Provenance for Concurrent Program Debugging and Analysis

### 2013-11-08 paper.tex
committing for sw update

\title{\Large Last Writer Slices and Communication Traps}

@@

\begin{abstract}
%Monitoring inter-thread memory dependences in production is useful.  Doing so
%improves debugging information and provides information to dynamic analyses and
%runtime systems.  Prior work that continuously monitors cross-thread
%dependences typically has overheads too high for production.

We design efficient system support for collecting {\em last writer slices} for
shared-memory concurrent programs.  Last writer slices dynamically track memory
updates.  Last writer slices reflect the provenance of values in memory,
information that is useful during debugging.   Using last writer slices,
we build {\em communication traps} (\ctraps).  \ctraps is a mechanism that
efficiently monitors and can interpose on memory operations that share data
between threads.  \ctraps is an extensible framework, exposing
inter-thread dependences to {\em \ctraps applications} that can implement
arbitrary dependence analyses.

@@

A common approach to this challenge is to reason about dependent memory
operations.  Memory operations are dependent if their order determines the
values they read or write.  The utility of dependence information is made
especially clear by its prevalent use during manual program debugging and in
support of program analysis tools.

Debugging a concurrent program requires understanding how threads interact and
dependence information furnishes that understanding.  Dependence information
helps programmers with the difficult task of reasoning about the {\em
provenance} of a memory location's value -- control- and data-flows that
influenced that value.  Programmers debug programs more effectively when they
have access to a value's provenance at points during a program's
execution~\cite{tipslicingsurvey,whylineicse}.  In concurrent programs, data
provenance information is particularly valuable because data may originate from
any of a potentially large number of threads, making reasoning without explicit
provenance information complex and difficult.

Compounding this complexity, many concurrency bugs manifest as failures rarely.
When a deployed system fails and a programmer receives a bug report,
reproducing the reported failure to collect provenance information may be
difficult or impossible.  It is essential to collect as much information as
performance margins permit from deployed systems, to reduce the need for
time-consuming failure reproduction.  Unfortunately, prior approaches to
tracking dependences in concurrent code~\cite{raceslicing,tipslicingsurvey}
impose high run time overheads.  These overheads are prohibitively high for use
in production.

@@

Dependence information is also an important part of many program analysis
tools.    Bug detection~\cite{avio,fasttrack,raceslicing,dmtracker} and
debugging~\cite{tipslicingsurvey,bugaboo,recon,cci,defuse,conseq,falcon} tools
analyze inter-thread dependences to find buggy behavior and help programmers
understand their errors.
Profilers~\cite{threadclustering,schedpredictionmodel} analyze inter-thread
dependences to find optimization opportunities.  Dependences are also essential
to some program understanding systems~\cite{oshatr} and dynamic checkers for
thread interaction properties~\cite{oshajava,velodrome}.
Record-Replay~\cite{chimera,doubleplay,fdr,rtr} systems track and record the
order of dependent events to help reproduce failures.  These systems would be
even more valuable, if their underlying dependence analysis was run in
production.  However, the heavy-weight analyses proposed by many of these
systems impose overheads too high for production use (5-100x
slowdown)~\cite{raceslicing,dmtracker,velodrome,recon,defuse,oshajava,chimera,coredet,stm}.

@@

Our goal in this work is to collect, with deployable efficiency, useful
dependence information that captures data provenance for use in debugging and
serves as a foundation for more sophisticated program analyses.  We start from
two simple observations: (1) Reads are frequent so we must minimize tracking
overheads on reads, and (2) In concurrency analyses, {\em inter-thread}
dependences are most important.  Based on these observations, we focus first on
tracking write-after-write dependences (WAWs), which provide a useful form of
data provenance information.  Building on that, we describe support for
tracking {\em communication dependences}.  Communication occurs when a thread
reads or overwrites a memory location that was most recently written by a
different thread; {\em i.e.}, inter-thread read-after-write dependences (RAWs)
as well as WAWs.  In this work, we demonstrate that this form of light-weight
dependence tracking is useful and versatile, and has inherent overheads that
are low enough for use in deployed systems.  There are two parts to our work:

First, we describe {\em last writer slices}
(Section~\ref{sec:lastwriterslices}).  Last writer slices record the thread and
program point that last wrote each memory location.     Last writer slices are
useful because they expose the provenance of a memory location's contents
during debugging (Section~\ref{sec:debugging}).  When a failure occurs, last
writer slices are saved with a core dump.  Last writer slices are also
available during an execution in the debugger ({\em e.g.}, at breakpoints).
Note that while prior work ({\em e.g.},
~\cite{recon,fasttrack,conmem,conseq,cci}) has relied on {\em ad hoc}
last-writer analyses, in this work, we isolate the essential last-writer
information in the form of last writer slices and provide an implementation
with overheads appropriate for production systems.  Tracking last writer slices
has low overhead because it imposes overhead on only write operations:  reads
are overhead-free.  In Section~\ref{sec:eval} we experimentally show that
collecting last writer slices indeed has overheads suitable for production
systems for a variety of applications.

Second, on top of last writer slices we build {\em Communication Traps} or {\em
\ctraps}  (Section~\ref{sec:ctraps}).  \ctraps monitors inter-thread
communication with overhead low enough for production use in many important
cases.  When threads communicate, \ctraps delivers a software {\em trap} that
can be handled by a trap handler.  Trap handlers implement {\em \ctraps
applications} through an extensible interface that can analyze or interpose on
the execution (Section~\ref{sec:apps}).  By default, \ctraps has very low
overhead.  \ctraps applications can perform arbitrary dependence analyses, but
at their own discretion and cost.  As we show in our evaluation, the overhead
of \ctraps applications scales with the complexity of their analysis.

>>>>

\title{\Large Last Writer Slicing: Efficient Data Provenance for Concurrent Program Debugging and Analysis}

@@

\begin{abstract}
%Monitoring inter-thread memory dependences in production is useful.  Doing so
%improves debugging information and provides information to dynamic analyses and
%runtime systems.  Prior work that continuously monitors cross-thread
%dependences typically has overheads too high for production.  
{\bf TODO Re-write}
We design efficient system support for collecting {\em last writer slices} for
shared-memory concurrent programs.  Last writer slices dynamically record the
code responsible for shared variable updates.  Last writer slices reflect the
provenance of values in memory, information that is useful during debugging.
Using last writer slices, we build {\em communication traps} (\ctraps).
\ctraps is a mechanism that efficiently monitors and can interpose on memory
operations that share data between threads.  \ctraps is an extensible
framework, exposing inter-thread dependences to {\em \ctraps applications} that
can implement arbitrary dependence analyses.

@@

%A common approach to this challenge is to reason about dependent memory
%operations.  Memory operations are dependent if their order determines the
%values they read or write.  The utility of dependence information is made
%especially clear by its prevalent use during manual program debugging and in
%support of program analysis tools.

Debugging a concurrent program requires understanding how threads interact and
{\em data provenance information} furnishes that understanding. A value's
provenance is a description of the reason it is held by a variable.  Provenance
can take on many forms, for example, a value's provenance could be the
instruction that wrote it or the function that returned it.  Provenance
information helps answer the question: "why does this variable contain this
value?".  During debugging a programmer can use that information to connect 
disparate parts of a program that interact, leading to a failure.

In concurrent programs, collecting debugging information, like data provenance,
from production systems is particularly important.  Concurrency bugs manifest
as failures rarely.  When a deployed system fails and a programmer receives a
bug report, reproducing the reported failure to collect provenance information
may be difficult or impossible.  It is essential to collect as much information
as performance margins permit from deployed systems, to reduce the need for
time-consuming failure reproduction.

Prior work has shown that provenence helps with debugging~\cite{badapples,
whylineicse, tipslicingsurvey}.  One effort tracked code that wrote unusable
values into variables~\cite{badapples}.  That work showed that revealing the
connection between a bad value's origin and a failure it causes is helpful
during debugging.  Other work showed that programmers debug programs more
effectively when they can explicitly ask "why" questions during
debugging~\cite{tipslicingsurvey,whylineicse}.

These efforts have shown the promise of provenance, but are not a panacea.
First, prior work~\cite{badapples} has focused on providing provenance
information for only some unusable {\em values}, like {\tt NULL}.  Limiting the
applicability of their technique to only some failures, like {\tt Null Pointer
Exceptions}.  Second, while prior work has shown that limited provenance
tracking for Java programs is possible with overheads low enough for
production~\cite{badapples}, a more general form of provenance tracking for C
and C++ programs has been elusive.  Prior approaches that use
slicing~\cite{thinslicing, tipslicingsurvey, whylineicse} or heavy-weight memory
checkers~\cite{badapples} impose overheads too high for production for C and
C++ programs. Third, no prior technique that we know of has collected
provenance information that is tailored to debugging concurrent programs with
overheads that are low enough for use in production systems.  Focusing on
concurrency is important because a value could have provenance in any of a
large number of threads, making reasoning about why a variable has a value
complex and difficult.

In that context, we design a novel form of in-production provenance tracking
for concurrent programs with three goals.   (1) To collect a general form of
provenance information, not limited to certain values.  (2) To provide
provenance information that helps understand how threads interact, which is
especially useful for concurrent program debugging.  (3) To work with overheads
low enough for use in production (ideally, less than 10\%).

We develop a novel provenance tracking mechanism called {\em last writer
slicing} that satisfies these goals.   As the program executes in production,
last writer slicing tracks the thread and program point that last wrote any
value to each potentially shared memory location --- its provenance.  Last
writer slices are saved with a core dump when the program fails and are
available in the debugger ({\em e.g.}, for breakpoint debugging).

Last writer slices are primarily useful for concurrency debugging, but we also
observe that provenance information in the form of last writer slices is a
useful foundation for implementing other concurrent program analyses.  Several
tools from prior work has used {\em ad hoc} last-writer information as part of
more heavy-weight, sophisticated in domains like bug detection ({\em e.g.},
~\cite{dmtracker,avio}), automatic debugging ({\em e.g.},
~\cite{recon,conmem,conseq,cci}), profiling ({\em
e.g.},~\cite{threadclustering, schedpredictionmodel}), and program
understanding~\cite{oshatr}.

Motivated by these prior {\em ad hoc} efforts, we build {\em Communication
Traps} or {\em \ctraps}  on top of last writer slices
(Section~\ref{sec:ctraps}).  \ctraps checks the last writer slice of each potentially
shared memory location when threads access those locations.  If the accessing thread
is different from the thread that last wrote, the access signfies {\em communication}
between the threads.  Communication is a common subject of

monitors inter-thread communication with
overhead low enough for production use in many important cases.  When threads
communicate, \ctraps delivers a software {\em trap} that can be handled by a
trap handler.  Trap handlers implement {\em \ctraps applications} through an
extensible interface that can analyze or interpose on the execution
(Section~\ref{sec:apps}).  By default, \ctraps has very low overhead.  \ctraps
applications can perform arbitrary dependence analyses, but at their own
discretion and cost.  As we show in our evaluation, the overhead of \ctraps
applications scales with the complexity of their analysis.

%Provenance is also useful in implementing concurrent program analyses.
%Dependence information is also an important part of many program analysis
%tools.    Bug detection~\cite{avio,fasttrack,raceslicing,dmtracker} and
%debugging~\cite{tipslicingsurvey,bugaboo,recon,cci,defuse,conseq,falcon} tools
%analyze inter-thread dependences to find buggy behavior and help programmers
%understand their errors.
%Profilers~\cite{threadclustering,schedpredictionmodel} analyze inter-thread
%dependences to find optimization opportunities.  Dependences are also essential
%to some program understanding systems~\cite{oshatr} and dynamic checkers for
%thread interaction properties~\cite{oshajava,velodrome}.
%Record-Replay~\cite{chimera,doubleplay,fdr,rtr} systems track and record the
%order of dependent events to help reproduce failures.  These systems would be
%even more valuable, if their underlying dependence analysis was run in
%production.  However, the heavy-weight analyses proposed by many of these
%systems impose overheads too high for production use (5-100x
%slowdown)~\cite{raceslicing,dmtracker,velodrome,recon,defuse,oshajava,chimera,coredet,stm}.
%

%First, we describe {\em last writer slices}
%(Section~\ref{sec:lastwriterslices}).       Last writer slices are
%useful because they expose the provenance of a memory location's contents
%during debugging (Section~\ref{sec:debugging}).  When a failure occurs, last
%writer slices are saved with a core dump.  Last writer slices are also
%available during an execution in the debugger ({\em e.g.}, at breakpoints).
%Note that while prior work ({\em e.g.},
%~\cite{recon,fasttrack,conmem,conseq,cci}) has relied on {\em ad hoc}
%last-writer analyses, in this work, we isolate the essential last-writer
%information in the form of last writer slices and provide an implementation
%with overheads appropriate for production systems.  Tracking last writer slices
%has low overhead because it imposes overhead on only write operations:  reads
%are overhead-free.  In Section~\ref{sec:eval} we experimentally show that
%collecting last writer slices indeed has overheads suitable for production
%systems for a variety of applications.

@@

@@

%Our goal in this work is to collect, with deployable efficiency, useful
%dependence information that captures data provenance for use in debugging and
%serves as a foundation for more sophisticated program analyses.  We start from
%two simple observations: (1) Reads are frequent so we must minimize tracking
%overheads on reads, and (2) In concurrency analyses, {\em inter-thread}
%dependences are most important.  Based on these observations, we focus first on
%tracking write-after-write dependences (WAWs), which provide a useful form of
%data provenance information.  Building on that, we describe support for
%tracking {\em communication dependences}.  Communication occurs when a thread
%reads or overwrites a memory location that was most recently written by a
%different thread; {\em i.e.}, inter-thread read-after-write dependences (RAWs)
%as well as WAWs.  In this work, we demonstrate that this form of light-weight
%dependence tracking is useful and versatile, and has inherent overheads that
%are low enough for use in deployed systems.  There are two parts to our work:
